---
title: "Delicacies"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```

Exercise
--------

### Missing data

We've already seen the missing data value: `NA`.  This special value allows R to handle missing data gracefully.  However,
just like in other programming languages you sometimes have to be careful how you handle `NA`.

```{r}
NA
is.na(NA) # why is.na vs. is.NA?  poor choice IMO.
is.na(2)
myvec <- c(1, 2, NA, 3)
is.na(myvec)
mean(myvec)
```

What happened?  Why is mean of `myvec` `NA`?  By default, the `mean` function does this intentionally so that you know that the 
`vector` is missing data because at some level it is unclear what you want to do.  Under most circumstances you just want
to remove the missing values and calculate the mean on what you have:

```{r}
mean(myvec, na.rm = TRUE)
```

This is also important when dealing with `data.frame`'s.  Let's inject some missing data into `esoph`.

```{r}
set.seed(596)
esona <- esoph
esona[sample(NROW(esona), 3), "ncases"] <- NA
esona[sample(NROW(esona), 3), "ncontrols"] <- NA
summary(esona)
```

Note there are now three `NA`'s tabulated by the `summary` function which are ignored in the summary statistics.  
However, if you try to take the mean of the `ncases` column without specifying `na.rm` you'll get `NA` again, just 
like above:

```{r}
mean(esona$ncases)
mean(esona$ncases, na.rm = TRUE)
```

You may be interested in which are and the number of complete cases:

```{r}
complete.cases(esona)
sum(complete.cases(esona))
# how many incomplete
sum(!complete.cases(esona))
```

You can also extract the `data.frame` where every row has complete data (i.e., the complete cases):
```{r}
esocc <- na.omit(esona)
NROW(esocc)
```

When you have missing data and you are subsetting you have to be more careful.  Let's look at the following:
```{r}
esona$ncases == 2
```

See the `NA`'s in there?  So, when you select those rows using the techniques in the ["Food Prep"](foodprep.html) module
there is some serious weirdness:
```{r}
esona[esona$ncases == 2, ]
```

Why?  Look at these examples:
```{r}
# a vector selection
x <- c("a", "b", "c")
x[c(NA, 2, 1)]

# back to the data frame
esona[c(NA, 2, 1), ]
```

So, when R finds an `NA` in a vector to be used as a slice it returns a empty version (in the `data.frame` case above an empty
row).  How do we work around that?  Incorporate the `is.NA` test:

```{r}
# says select ncases that are equal to 2 and not NA
esona[esona$ncases == 2 & !is.na(esona$ncases),]
```

That works because anything that is `NA` will be set to `FALSE` instead of `NA` and not included in the selection.
Subset does not have this problem, but remember you cannot assign directly to a `subset` so you'll need the above
syntax sometimes.

```{r}
subset(esona, ncases == 2)
```

Why?  Because if you read the help for subset,it says uses similar syntax internally to that we used above!

### Other special values

You met a lot of these in the very [first module](rithmatic.html):

```{r}
5/0
-5/0
0/0
```

`NaN` means "not a number", `Inf` means "positive infinity", and `-Inf` means "negative infinity".  They are almost always seen because you are dividing by zero, usually unintentionally.  You can test for these values 
which can help you find bugs in your programs.

```{r}
is.nan(0/0)
is.infinite(1/0)
is.finite(1/0)
```

Finally, the `NULL` value is a marker for "nothing."  Not nothing as in something that ought to be there but is not (e.g., missing), but literally, something that does not or should not exist in the realm of R.  It is a good way to remove a column from a `data.frame`:

```{r}
esocc$ncases <- NULL
head(esocc)
```

### Other useful set and subsetting functions

`duplicated` finds duplicated values.  It also works to find duplicated rows in
`data.frames`.  `unique` will extract the values that are, well, unique!

```{r}
x <- c(4, 8, 0, 1, 8, 9, 10, 4, 2)
duplicated(x)
unique(x)
```

These functions allow you to make selections based on "set" operations:
`union` joins two things together (without duplication), `intersection` finds
the common values between two vectors, and `setdiff` finds the difference 
between two vectors.

```{r}
union(c("a", "b", "c"), c("b", "c", "d", "f"))
intersect(c("a", "b", "c"), c("b", "c", "d"))
setdiff(c(1, 2, 3, 8, 9), c(2, 8))
```

### Text

#### Pasting

We have seen `paste0` which pastes a series of strings together without any other 
character between.

```{r}
paste0(c("a", "b", "c", "d"), 1, c(1, 2))
```

The more general function is `paste`.  It takes two options: `sep` which is
the separator between the elements.  It can also take a `collapse` argument
which if set pastes the entire sequence of strings together using that 
character.

```{r}
paste(c("a", "b", "c"), 1, c(1, 2), sep = " ")
paste(c("a", "b", "c"), 1, c(1, 2), sep = " ", collapse = "-")
```

#### Substrings

Sometimes you need to select out a piece of a string.  The `substr` function
is helpful for fixed length strings.  Note that the `[1:3]` select out the 
first three elements of the names of esoph while the other 1 and 3 tell it to
take the first through the third of each string.  They are NOT related.

```{r}
names(esoph)[1:3]
substr(names(esoph)[1:3], 1, 3)
```


#### Regular expressions / Searching

**Regular expressions** are a very powerful way of finding patterns in text and
manipulating them.  You could take an entire course on **regular expressions**
so I will only be able to introduce you to some basic aspects.  

Let's start with a list of words:
```{r}
words <- c("art", "bat", "bet", "bee", "bees", "beet", "believe", "beat", 
           "cat", "car", "cars", "can", "cart", "mississipi",
           "mart", "part", "see", "sat", "set")
```

We will use the `grep` function which takes a **regular expression** and 
character vector to search on.  If you just use a sequence of letters or numbers
you will find strings that match that pattern of letters and numbers:

```{r}
grep("s", words)
# to see the matched words
grep("s", words, value = TRUE)
```

Note how each contains an "s".  Here is another example:
```{r}
grep("ee", words, value = TRUE)
```

Nothing ground breaking there... however there are characters with special
meanings.  Such as "." that will match any character.  To find words that have
an "a" followed by any character:
```{r}
grep("a.", words, value = TRUE)
```


```{r}
grep("s.t", words, value = TRUE)
```

You can put a group of characters in square brackets:
```{r}
grep("[bs].t", words, value = TRUE)
```

The `*` character matches zero or more characters matching the previous expression
(character or group):
```{r}
grep("beliea*ve", words, value = TRUE) 
```

The `+` character matches one or more of the previous expression:
```{r}
grep("beliea+ve", words, value = TRUE) # nothing matches
grep("be+t", words, value = TRUE)
```

Finally you can group things in parentheses:
```{r}
grep("m(iss)+ipi", words, value = TRUE)
```

I find regular expressions to be very helpful for selecting a series of variables
in a `data.frame`:

```{r}
grep("gp", names(esoph), value = TRUE)
head(esoph[, grep("gp", names(esoph), value = TRUE)])
```

The `grepl` function is also useful as it gives the `logical` vector of whether
a match is found in a sequence an may be useful in some settings where the index
or value will not work:

```{r}
grepl("gp", names(esoph))
```

There is so much more you could learn about this, but even this basic introduction
can take you very far.  For more information, I recommend you check out this
online tutorial: <http://www.zytrax.com/tech/web/regex.htm>.

Please note that you can use regular expressions to select out parts of text,
rearrange, or manipulate various strings.  See the `sub` and `gsub` functions for
more information.



### Dates

Dates are often difficult to work with because there are many different formats and methods of converting
to a system that the computer can work with.

In R, the default date format is `YYYY-MM-DD`, i.e., 4 digit year, hyphen, 1–2 digit month, hyphen,
and 1–2 digit day. The as.Date() function allows you to take character strings and convert them to dates
that the computer can understand.

```{r}
as.Date("2011-01-12")
```

You can also use other formats if you receive data from someone who typed it in differently by providing
a format argument:

```{r}
as.Date("01/12/2011", format="%m/%d/%Y")
as.Date("12jan2011", "%d%b%Y")
```

For more information, read the help file for `strptime`. The `difftime` function allows you calculate various intervals
between dates:

```{r}
difftime(as.Date("2011-1-14"),as.Date("2011-1-12"))
difftime(as.Date("2011-1-14"),as.Date("2011-1-12"),units="hours")
```

Explore and Extend
------------------

+ From the `esona` `data.frame` created above, select the rows where `ncases`
equals `3` but not `NA` (without using `subset`).

+ For regular expressions, the `^` is used to represent the beginning of a 
string and `$` is used to select the end of a string.  Use these to select
the words starting with `b` from `words` above.  Next, try to get only those
words that end with `s` using one of them.

+ Presuming that the string `"12Oct1994"` represents a 
date that is the same as October 12, 1994
read the appropriate help file discussed above and convert that
string to a date.

+ How many days have you been alive?

Evaluate
--------

Pick a data set of your own choosing (perhaps this 
is an opportunity to test drive the data you want to use for your
final project?).  Use one of the text or date functions to do something 
you feel is useful to your data.  Explain why and demonstrate it in a knitted `.Rmd` file.
