---
title: "Functions"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```

Exercise
--------

### The anatomy of a function

Data are the "nouns" of a computer language while functions are the "verbs."  I also like to think of functions 
as a type of creature that eats and processes
data thereby providing a useful output.  Think of yeast which "eat" sugar and process it into alcohol.  
Functions have a name, a mouth, and a body.  Take the
`c` function as an example.  The name of the function is `c`.  The mouth is the parentheses.  

```{r}
c("e", "a", "t")
```

The food called **arguments** goes in the mouth separated by commas (I like to think of them as teeth; OK now, maybe I'm taking the analogy too far...).  

Where is the body?  R functions are shy and do not like to show off their body.  They hide it, but if you call their name
without offering food, they will show it to you as they turn to run to run away realizing you do not
plan to offer them anything to eat.

```{r}
c
```

`c` has an unusually strange body.  Dark magic is going on there so I suggest you just look away, unless you are into that sort of thing.  Let's look at a more typical function, `factor`, which you met in [Rcel, part 1](rcel1.html)

```{r}
factor
```

There is a lot going on here, but look at the general structure.  The word `function` followed by the mouth enclosed in parentheses `()` telling
you some details about the foods this creature eats.  Next, you see a section enclosed in curly brackets `{}`: that's the body.  What is inside?
R code!  Basically, inside the body of a function is a recipe of R code that can be reused to repeat a useful action.  

If you thought `c` was odd, there are even functions that do not show their mouth without some coaxing.  Can you think of one?  What is an
"action" you have taken on data that did not need parentheses?  Think about for a moment before continuing.

How about the **operators**?

```{r}
2 * 2
2 * c(1, 25, 100)
```

Is `*` a function?  Indeed, it is, but you have to know how to get these painfully shy creatures to show their mouth.  To do so, you need to
get bold with calling their name by using backticks. (Remember those? Under the tilde, `~`, on the left side of your keyboard usually.)

```{r}
`*`(2, 2)
`*`(2, c(1, 25, 100))
`*` # show *'s body, more dark magic
```

### Default and named arguments

All arguments have names, but you do not always need to use them.  Look at the help for
`factor`

```{r, eval = FALSE}
?factor
```

You'll see the following near the top:
```{r, eval = FALSE}
factor(x = character(), levels, labels = levels,
       exclude = NA, ordered = is.ordered(x), nmax = NA)
```

In order the names of the arguments for
the `factor` function are `x`, `levels`, `labels`, `exclude`, `ordered`, and `nmax`.  After
many you will note an equals sign with something after it.  These are the **defaults** for
the argument if you do not specify them.  However, just because `levels` does not have 
an explicit **default** in the list does not mean that it does not have one.  If you read further
into the help it says that `levels` does have a **default** value.  

Since all values have a **default**, the function `factor` will return valid output even if run with
no arguments:
```{r}
factor()
```

On the other hand `+` does not have **default** arguments:
```{r, error = TRUE}
`+`()
```

So why the names?  The most important reason as you will learn in the next section is so
that you can refer to the arguments by name in the body of the function.  A secondary use has
to do with the way you call functions.  If you do not name the arguments when you call the function,
R assumes that you are using them in order starting with the first.  Names are useful when
you need to skip an argument, want to be explicit about which argument something is, or if
you might be using them out of order:

```
a <- c(1, 1, 2, 2)
factor(a, labels = c("M", "F"))      # skipping levels argument
factor(x = a, labels = c("M", "F"))  # being explicit
factor(labels = c("M", "F"), x = a)  # out of order
```

### Writing your own functions

Functions are your friend.  Have you ever cut-and-paste a piece code to run it again
after just tweaking it to run on slightly different data.  Try not to do that, instead use
functions.  Why?

+ When you use functions, you encapsulate little pieces of logic in your program.  It makes your
code and the thinking behind it much easier.  The code will also be shorter overall.
+ If you cut-and-paste and realize you need to change something you have to do it everywhere and
not make any mistakes, with 
a function you do it in one place: the function.
+ You will likely find your functions are useful in your next project and a function is a good way to
carry something useful from one project to another in a useful way.

A major principle of programming is DRY which stands for "Don't Repeat Yourself".  I might cut and paste
something twice but by the 3rd paste unless I'm seriously not going to do it again and it is a 
*very* short (one line) piece of code, I am going to be trying to make a function out of it.  This
rule of not using the same code more than three times is also known as the "rule of three" in
computer programming.

So how to write a function?  Name it (via assignment with `<-`), use the keyword `function` followed
by the mouth with a list of arguments.  Then, write the R code as the body.  Use it just like any
other R function.  Use the name you chose for the argument in the function to work on the data.
Examples:

```{r}
add2 <- function(x) { x + 2 }
add2(2)
add2(5)
```

```{r}
add2sub5 <- function(x) { 
  x <- x + 2
  x - 5
}
add2sub5(10)
```

The second example shows that within the body you need to write R code just like you would at the console
with each **statement** on a separate line.  R returns the result of the last statement.  The variables
defined in the function are **local** changing them as we did in `add2sub5` does not change them outside
the function.  This also makes your programs much safer (so you don't accidentally change something
you did earlier in a long program).

```{r}
x <- 10
add2sub5(10)  # inside makes x == 12
x             # once it returns x is still 10
```

However, if you use a **uninitialized** variable within your function R can use the one from the
environment:

```{r}
y <- 10
addysub5 <- function(x) {
  x <- x + y
  x - 5
}
addysub5(10)

y <- 2
addysub5(10)
```

### A note on naming
Remember when we said that you can not use one of the **reserved** words for a **name**.  If not, go back to the [Rcel](rcel.html) exercise.  
However, you also need to be careful with things like `c` even though they are *not* reserved words when naming functions.  Try the following:

```{r}
c <- function(x) { x + 5 }
```

```{r, error=TRUE}
c(1, 2, 3)     # Uh oh.
```

You have hidden the usual `c` with your `c`: 
```{r}
c(2)
c
```

So, get rid of it with the `rm()` function (rm is the mnemonic for remove):
```{r}
rm(c)
c(2) 
c(1, 2, 3)
```

However, R treats data differently than functions so
named data generally does not conflict with other functions
that you normally have access to.  It is still obviously confusing
and bad practice to do this (plus you cannot create a function and a piece
of data named the same thing anyway from the console or code without 
some serious voodoo - one would replace the other:

```{r}
c <- c(1, 2, 3)
c(2, 3, 4)
c(c, c)
c <- function(x) { x + 5 }
c(2)
rm(c)
c(2)
```

Explore and Extend
------------------

+ Add two numbers together using the "mouth" of the `+` function.

+ Write a function called `add5` that adds five to a single number.

+ Write a function that takes two arguments a `vector` and a single number and returns the element of the
vector corresponding to that number.  Call it `pick_n`.

+ Write a function that computes the two values that solve  
a quadratic equation of the form, $ax^2 + bx + c = 0$.  The values as you
may recall are called roots are given by the following 
formula that you've probably not seen since high school:
$$ {-b \pm \sqrt{b^2 - 4ac}}\over{2a} $$
You can compute the square root with `sqrt()`.  R only returns the last object... how can
you return "two" things at once?  (Hint: use one of the data types you already know.)

+ Try to write a function that returns a function.  Here is a suggestion.  Write a function that
takes a single number as an argument and returns a function that adds that number to its argument.
Remember you can save the output of a function like anything else.  So, your function, if it were
called `add_factory`, would work like this:

```{r, echo = FALSE}
add_factory <- function(x) { function(y) y+x }
```
```{r}
add6 <- add_factory(6)
add6(6)
add10 <- add_factory(10)
add10(6)
```

Evaluate
--------

1. Look at the `gl` function.  Give a working example where you use the `n`, `k`, and `length` arguments out of the usual order.
```{r, opts.label='hideall'}
gl(length = 20, k = 1, n = 2)
```

2. Write a function that [cubes](https://en.wikipedia.org/wiki/Cube_%28algebra%29) numbers.  Call it `cube`.
```{r, opts.label='hideall'}
cube <- function(x) { x^3 }
```

3. Write a function that takes a `character` name of one of the columns in a `data.frame` and a `data.frame` and returns that column from a `data.frame`. Call the
function `select_column`. (You don't really need a function for this, so don't
over think it, but it is good practice.)
```{r, opts.label='hideall'}
select_column <- function(name, df) { df[, name] }
```

4. The `length` function gives you the length of vector and `sum` adds a numeric vector together.  Write a function called `mymean` that takes a vector (of arbitrary length) as input and calculates the average of the numbers in that vector using the `length` and `sum` functions.
```{r, opts.label='hideall'}
mymean <- function(x) { sum(x) / length(x) }
```

5. Using your `mymean` function write a function that calculates the variance called 
`myvar`.  Remember variance is the sum ($\sum$)
of the squares of values after you subtract the mean ($\bar x$) of all the values ($x_i$) from each divide by the length ($n$)
minus 1.
$$ var = {{\sum(x_i - \bar x)^2}\over{n-1}}$$
```{r, opts.label='hideall'}
myvar <- function(x) { sum((x - mymean(x))^2) / (length(x) - 1) }
```