---
title: "Getting Data In & Out of R"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```

Prepare
-------

This is not a typical module.  There is only a brief tutorial on getting datasets in and out of R as well as some
file system related commands.

Excel files
-----------

Next, install the package `readxl` which allows you to read Excel files:

```{r, eval = FALSE}
install.packages('readxl')  # only need to do this once per installed version of R
```

To read an Excel file into R you load the `xlsx` library once and use the `read.xlsx` function. 
```{r}
library(readxl)  # only once per session
```

 Because I've saved the files we are going to
use in the `adm` package you need to ask R where the package directory is (your answer will be different 
depending on your operating system and what version of R you are running, etc.).  These steps are
 unique for accessing a raw data file contained in a package.  You will NOT
 generally do this, but it is a convenient way to distribute the file to you.  
 Once you discover where the file is however, the rest of the steps are the same.
```{r}
system.file(package = "adm")
```

The Excel file we are going to use is a version of the `esoph` dataset that is in the `extdata` directory of the package.  If you get
a blank `character`, you have typed something wrong or you `adm` package is not up-to-date.
```{r}
system.file("extdata/esoph.xlsx", package = "adm")
```

With our filename in hand, we can now call the `read.xlsx` function.  The `1` tells it to pull out the first sheet of the file:
```{r}
read_excel(system.file("extdata/esoph.xlsx", package = "adm"), 1)
```

Hopefully, it looks familiar.  Of course you can save it for later:
```{r}
xesoph <- read_excel(system.file("extdata/esoph.xlsx", package = "adm"), 1)
```

`read_excel` automatically reads in the `data.frame` as though you called `data.frame` with
`stringsAsFactors = FALSE`.  In general this is the best way to import
any data due to the problems you have seen when you start with `factor`
and want to do additional data manipulations.  
We will soon learn ways to convert variables to factors
 nearly automatically anyway.

Comma Delimited Text Files
--------------------------

Comma delimited text or comma separated values (CSV) are a common text file format for representing data.  You can use the 
builtin `read.csv` to read CSV files:
```{r}
iih <- read.csv(system.file("extdata/iih.csv", package = "adm"), stringsAsFactors = FALSE)
```

Excel can open and write CSV too.

File system commands
--------------------

R is working in a directory of your file system.  You can figure out which one using `getwd` which means "get working directory":
```{r, eval = FALSE}
getwd()
```

You can change this directory using `setwd`:
```{r, eval = FALSE}
setwd("/Users/bbbruce/Documents")  # you need to modify this for your system and preferences
```

Since you may not be familiar with your file system, I recommend using RStudio's menu.  Go to
Session -> Set Working Directory -> Choose Directory... so that you can browse to where you want to 
work.  When you are in the working directory, you do not need any of the fancy commands we were using earlier to
figure out where the file is.

For example, if you had a file called "iih.xlsx" in your working directory you would read it in like this:
```{r, eval = FALSE}
read_excel("iih.xlsx", 1, stringsAsFactors = FALSE)
```

Likewise, if you write anything out, it will be in the working directory unless you tell R otherwise.  So if you cannot find something
you need make sure to check what directory you were working in!

Writing data files
------------------

There are write functions for CSV (which you can open in Excel easily):
```{r, eval = FALSE}
write.csv(iih, "iih2.csv")
```

These will be written into your working directory.  Be careful, if they exist, they will be overwritten.

You can also directly save R objects:
```{r, eval = FALSE}
saveRDS(xesoph, file = "xesoph.rds")   
saveRDS(iih, file = "iih.rds")  

rm(xesoph)
rm(iih)

xesoph <- readRDS("xesoph.rds")      
iih <- readRDS("iih.rds")  
```

I often use these functions in conjunction with `file.exists` to cache long data operations so that I don't have to wait
for them to finish everytime I run my code.

```{r, eval = FALSE}
if(!file.exists("iih.rds")) {
  # ... do time consuming task to create iih
  saveRDS(iih, file = "iih.rds") 
} else {
  iih <- loadRDS("iih.rds")
}
```