---
title: "Rcel, part 2"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```

Exercise
--------
In the [first part of this exercise](rcel1.html) we learned about `vector` and `factor` which are two 
building blocks of a `data.frame`.  As you recall, a `vector` requires that each element be of the same
**atomic** data type.  Now, we meet the first **composite** data type that can contain multiple **atomic** 
types as elements simultaneously: `list`.  A specialized `list` called `data.frame` will be used to store our datasets and 
will allow us to operate on our datasets as spreadsheet-like objects.

## Lists
A `list` allows you to mix 'n' match atomic data types.  You bulid a list with the function `list()`.  

```{r}
l <- list("A", 2, TRUE)
is.character(l[[1]])  # see the subtle difference between l ("ell") and 1 ("one")?
is.numeric(l[[2]])
is.logical(l[[3]])
is.list(l)
```

Did you notice the `[[]]` syntax used to index a `list`?  So, each element of a `list` does not have to be just a single atomic element, but can be a `vector`:

```{r}
l <- list("A", c(1,3,2,4), TRUE)
l # notice the hints that R gives you about how to access a given element like it did for vectors?
l[[2]]
```

You can also name the elements, and using the `$` syntax you can access those elements using that name:

```{r}
l2 <- list(a = "A", b = c(1, 3, 2, 4), c = TRUE)
l2 # again notice R's hints about accessing elements
l2$c
```

## Extending assigment

With a `vector` you can **reassign** a specific element like this:

```{r}
x <- c(1, 3, 2, 4)
x[2] <- 2
x
```

You can also assign to parts of a `list` using the list syntax (see Explore and Extend below).

## Data frames

A `data.frame` is a specialized `list` where all the elements of the `list` have equal length.  It is perfect for
representing data where there are several values (in columns) per observation (in rows).  You create a `data.frame` with the `data.frame()` function:

```{r}
df <- data.frame(age = c(3, 2, 3, 3, 1, 2, 4, 4), 
                 sex = factor(c("m", "f", "m", "f", "m", "f", "m", "f")), 
                 sorethroat = factor(c("no", "no", "no", "no", "yes", "yes", "yes", "yes")))
df
df$sex
summary(df)
```

In general, we will usually import a `data.frame` from an external file, like an Excel spreadsheet, but there are times when you want to build
a `data.frame` from scratch, especially as you attempt to restructure data.

You can check what variables are in a `data.frame` like this (it also works for `list`):

```{r}
names(df)
```

We can reference an exact row and column pair in a `data.frame` (rows first, columns second - this is the usual convention in mathematics and
programming):

```{r}
df[2, 3]
```

If we leave out the row or column specification, we get the whole row or column:
```{r}
df[2, ] # entire second row
df[, 3] # entire third column
```

You can use the name of the column too:
```{r}
df[2, "sorethroat"]  # same as df[2, 3]
df[, "sorethroat"]   # same as df[, 3]
```

Finally, you can use the `subset` command to find rows that match certain criteria:
```{r}
subset(df, age == 2)
subset(df, age == 2 & sorethroat == "yes")
```

This is a new `data.frame` that you can subset and operate upon:
```{r}
subset(df, sorethroat == "yes")$age
median(subset(df, sorethroat == "yes")$age)
```

Explore and Extend
------------------

+ Demostrate two ways to extract the numeric element `2` from the second element of `l2` above. (Hint: combine the
syntax from this exercise with that of the last one and don't forget to look for
 the subtle difference between l ("ell") and 1 ("one") which I'm doing to keep 
 you on your toes - that's not `12` but `l2`).

+ What data type is `l2[1]` vs. `l2[[1]]` vs. `l2$a`?  What happens when you do `l2[c(1,2)]` vs. `l2[[c(1,2)]]`?  (Notice the
difference between this and how we reference rows and columns of a `data.frame`.) You know that you can extract
elements from a `vector` using `[]`. So now, can you extract a numeric vector equivalent to `c(3,4)` from `l2` by generalizing the
ideas from this question and the last one?

+ Demostrate two ways to reassign the second element of `l2` to a value of `900`. Notice that there is also a difference between `O` ("oh") and `0` ("zero").

+ Prove that a `data.frame` is a `list`.  Prove that an arbitrary `list` like `l2` is not a `data.frame` (guess what the function is to test if
something is a `data.frame`).  

+ In `df`, replace the 5th observation's sex with `"f"`.  Now, try to replace the 5th observation's sore throat with `"maybe"`.  Can you make sense
of the error message?  Can you create a solution such that at the end `df` still contains a factor called `sorethroat` with the help of the function `as.character`?  (By the way, there are `as.` versions of all the types we've studied.) It is used like this:

```{r}
as.character(df$sorethroat)
```

Evaluate
--------

R includes several datasets.  You can see what is included by running the `data()` command.  For this evaluation, we will use the `esoph` dataset.  You can
load it like this `data(esoph)`.  After that, you can refer to it with just `esoph`.  If you are curious about the data, run `?esoph` which will pull up the help file for the data.  This file is in a case-control format where each row represents multiple cases and controls, in variables `ncases` and `ncontrols`.
So, if you are asked how many observations there are you'd add the number of cases and controls.  Include the R code for each problem.  Do not do only by hand.

1. Display a simple summary of `esoph`.
```{r, opts.label='hideall'}
summary(esoph)
```

2. Extract the 4th row of `esoph`.
```{r, opts.label='hideall'}
esoph[4, ]
```

3. List the names of `esoph`.
```{r, opts.label='hideall'}
names(esoph)
```

4. What is the number of cases which have tobacco consumption (`tobgp`) of 30+ gm/day?  You'll find the command `sum()` helpful.  Use it like `median()` above. 

```{r, opts.label='hideall'}
sum(subset(esoph, tobgp == "30+")$ncases)
```

5. What is the number of *observations* where both the age group (`agegp`) is 25-34 years and tobacco consumption is 10-19 gm/day?
```{r, opts.label='hideall'}
sum(subset(esoph, agegp == "25-34" & tobgp == "10-19")[, c("ncases", "ncontrols")])

# another way of many
esoph.ss <- subset(esoph, agegp == "25-34" & tobgp == "10-19")
sum(esoph.ss$ncases) + sum(esoph.ss$ncontrols)
```
