---
title: "Branching Out & Feeling Loopy"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```

Exercise
--------

In the [last lesson](function.html) we had an introduction to functions, a way to encapsulate pieces of R code in a way that we substitute different values 
(**arguments**) into that code and apply that "recipe" to a set of arbitrary values.  However, what if we need the program to "think", i.e., to make decisions
about what to do?  What if we need to do something many, many times: ten times, one hundred times, a million times?  Yes, our function might be easier to write
than the code it runs, but you still don't want to write your function out that many times.  This lesson addresses both of these problems.

### Branch Control Structures

#### `If`-`else`

**Control structures** guide the flow of of your program.  The first types of **control structure** we will consider are **branch structures**.  They are
called **branch structures** because they place a "fork in the road" in your program.  The first one to look at is `if`:

```{r}
x <- 5

if (x > 3) {
  print("x is greater than 3")
}

if (x > 10) {
  print("x is greater than 10")
}
```

As you can see in the first case R gets to the word `if`, checks the condition in the parentheses, and if it is `TRUE` it runs the code inside the curly
brackets.  If the expression in the parentheses is `FALSE`, it skips the code in the brackets.  The `print()` function would print out the message that
is provided to it.  This isn't that important at the interactive console, but will be more important as you advance to running more complex programs.
You can add an `else` to provide an alternative block of code to execute if the expression is not `TRUE`:

```{r}
if (x > 10) {
  print("x is greater than 10")
} else {
  print("x is less than or equal to 10")
}
```

You can chain together a series of `if` and `else` statements.
```{r}
if (x > 10) {
  print("x is greater than 10")
} else if (x > 4) {
  print("x is greater than 4")
} else {
  print("x is less than or equal to 4")
}
```

You can nest `if`-`else` constructions:
```{r}
if (x > 3) {
  if (x > 4) {
    print("x is greater than 4")
  }
}
```

#### `Switch`

What if there are several choices?  You can use chained and nested `if` and `else` expressions, but often a simpler solution is the `switch` statement:

```{r}
center <- function(x, type) {
  switch(type,
         mean = mean(x),
         median = median(x),
         trimmed = mean(x, trim = 0.1))
}

x <- c(5, 0, 6, 8, 9, 7, 3, 4, 4, 7)

center(x, "mean")
center(x, "median")
center(x, "trim")
```

Like the builtin R functions you can provide defaults within the **argument** list:

```{r}
center <- function(x, type = "mean") {
  switch(type,
         mean = mean(x),
         median = median(x),
         trimmed = mean(x, trim = 0.1))
}
center(x)
```

Within the options you can include more complex code by wrapping it in curly brackets:

```{r}
center <- function(x, type = "mean") {
  switch(type,
         mean = mean(x),
         median = median(x),
         trimmed = mean(x, trim = 0.1),
         mode = {
           x <- factor(x)
           s <- summary(x)      # summarize the factor (table of how many each)
           m <- max(s)          # find the most frequent
           l <- levels(x)       # extract the levels
           o <- l[s == m]       # take out the the levels that match the max in the summary
           as.numeric(o)        # convert back to numbers, factor level names are characters
         })
}
center(x, "mode")
```

#### `ifelse`

`ifelse` is a function that provides a simple branch logic for a `vector`.  It is technically *not* a **control structure** because it does not
change the flow of your program, but it is very useful and similar in concept to `if`-`else`.

```{r}
x
ifelse(x == 7, "seven", "not seven")
ifelse(x %% 2 == 0, "even", "odd")
```

A new operator worth knowing is the `%in%` operator.  You will find the operator
is useful in many situations and can use it in other places than just inside
an `ifelse` function call:

```{r}
ifelse(x %in% c(7, 4, 0), "in", "out")
```


### Loop control structures

#### `for`

The `for` **loop control structure** allows you to **iterate** over a `vector` or `list`.  Here is an example:

```{r}
for (x in 10:1) {
  print(x)
}
```

Read that as `for` each element in the vector `10:1` (which recall is equivalent to `c(10, 9, 8, 7, 6, 5, 4, 3, 2, 1)`) call it `x` temporarily
and do what is inside the curly brackets.  You could also **iterate** over the elements of the object directly.  For example:
```{r}
l <- list(a = c(3, 4, 5), b = c("dogs", "cats"))
for (x in l) {
  print(x[1])
}
```

#### `while`

The `while` loop keeps doing something while (get it?) the condition of the loop is `TRUE`.  For example:

```{r}
i <- 10
while(i > 0) {
  print(i)
  i <- i - 1
}
```

If you aren't careful, you could write the loop wrong in a way that it would never end:

```{r, eval = FALSE}
i <- 10
while(i > 0) {
  print(i)
  i <- i + 1
}
```

This is an **infinite loop**.  To get R to stop, hit the little stop sign at the upper right of your console window (usually the lower
left panel of RStudio).

#### `break` and `next`

Two **reserved** words permit further control of loops: `break` and `next`.  When used within a loop (either `for` or `while`), they
allow you to completely break out of the loop and continue further along your program (`break`) or skip the rest of the current loop 
and start the next one (`next`).  They can help with trick situations.  Here are a couple of examples:

```{r}
i <- 10
while(TRUE) {
  if(i == 0) break;
  print(i)
  i <- i - 1
}
```

```{r}
i <- 20
while(TRUE) {
  if(i < 5) break;
  if(i %% 2) { 
    i <- i - 1 
    next; # go to next iteration if even
  }
  print(i)
  i <- i - 1
}
```

Now the latter is unnecessarily complex, but demonstrates the concepts. You will plan to make it simpler in "Explore and Extend".

Explore and Extend
------------------
+ Sometimes you want to get access to the index of a row or column of a `data.frame` or the index of an element in a vector or list inside
your `for` loop.  The functions `NCOL`, `NROW`, `length`, and `seq_along` can be a help.  Try each with `esoph` (remember this is a builtin
dataset in R).  Why is the length of `esoph` 5? 

+ Above we created a `while` loop using both `break` and `next`.  Rewrite the `while` loop using neither `break` nor `next`.

+ Recall that the `esoph` dataset is written in case-control format such that the table is compressed with the first three columns
representing patterns of covariates and the last two columns reporting the number of cases and controls with that pattern of covariates.
Some statistical methods will accept data in this format, but not all.  Sometimes you need to convert the dataset to "long" form where there
is one row for each subject and four columns.  The first three columns representing the pattern of covariates and the last, which we will call `case`
will be `case` if the subject was a case and a `ctrl` if the subject was a control.  Our ultimate goal is to convert the `esoph` `data.frame`
from its current format into the "long" format with those four variables as `factors`.

Let me give you some additional insight into the problem.  Start by examining row 72.  Remember how to do that?
```{r}
esoph[72, ]
```

We see that this pattern of covariates has 2 cases and 3 controls, so in the "long"" dataset we should have the following 5 rows. 
(Note: that they neither need to be contiguous nor in any particular order, but we need them somewhere in the new `data.frame`.)
```{r, echo = FALSE}
suppressWarnings(cbind(esoph[72, 1:3], case = c(1, 1, 0, 0, 0)))
```

Since manipulating factors gets a little tricky, I'd start the program by converting any `factor` to a `character`. (You learned how 
to do that last week.)

You can just use `c`  to append a single item to a `vector` or `list`.  For lists, if you try to add a more complex item, `c` does not
work and you need a more complex solution (like the one for `data.frame` below).
```{r}
l <- esoph[72, ]
is.list(l)
c(l, case = 1) 
```

To start a `data.frame` from scratch in this scenario is a little odd.  We'll learn better techniques later, but here is how
I would first create the new `data.frame` you are going to make. `stringsAsFactors = FALSE` tells R the `data.frame` function
not to convert any `character` vector into a `factor`: 
```{r}
new_df <- data.frame(agegp = "", alcgp = "", tobgp = "", case = 0, stringsAsFactors = FALSE)
```


Finally, you need to know "one weird trick" to append a row or column to a `data.frame` 
If I have the following `data.frame` and I want to add a row, I just assign to it as though it was there:
```{r, echo = FALSE}
df <- as.data.frame(lapply(esoph[1:5, 1:3], as.character), stringsAsFactors = FALSE)
```

```{r}
df
df[6, ] <- list(agegp = "35+", alcgp = "40-79", tobgp = "10-19")
df
```

Now, you have all the pieces to write the code to do this.
But let's not get ahead of ourselves.  The first step to a problem like this is to carefully think through the process before you
even begin to write a line of code.  We will keep breaking down the process until we can write code for each step, preferably 
encapsulating steps in functions where appropriate.  

```{r, opts.label = 'hideall'}
make_new_rows <- function(row) {
  ncases <- row$ncases
  ncontrols <- row$ncontrols
  out <- list()
  if (ncases != 0) {
    for (i in 1:ncases) {
      out[[i]] <- c(row[1:3], case = "case")
    }
  }
  if (ncontrols != 0) {
    for (i in (ncases + 1):(ncases + ncontrols)) {
      out[[i]] <- c(row[1:3], case = "cntl")
    }
  }
  out
}

# prepare the old data.frame
old_df <- esoph
old_df$agegp <- as.character(old_df$agegp)
old_df$alcgp <- as.character(old_df$alcgp)
old_df$tobgp <- as.character(old_df$tobgp)

# prepare a new data.frame
new_df <- data.frame(agegp = "", alcgp = "", tobgp = "", case = 0, stringsAsFactors = FALSE)

j <- 1 # to hold the row index in the new data.frame
for(i in 1:NROW(old_df)) {
    newrows <- make_new_rows(old_df[i, ])
    for(r in newrows) {
      new_df[j, ] <- r
      j <- j + 1
    }
}

new_df$agegp <- factor(new_df$agegp)
new_df$alcgp <- factor(new_df$alcgp)
new_df$tobgp <- factor(new_df$tobgp)
new_df$case <- factor(new_df$case)
```


Evaluate
--------

Write a program that converts the "long" format back into the case-control format.   

Start here:
```{r}
new_old_df <- unique(new_df[, 1:3])
new_old_df$agegp <- as.character(new_old_df$agegp)
new_old_df$alcgp <- as.character(new_old_df$alcgp)
new_old_df$tobgp <- as.character(new_old_df$tobgp)
new_old_df$ncases <- 0
new_old_df$ncontrols <- 0
```

Use the following code fragment plus the techniques in this lecture to finish the job.  Do not forget that you can assign to these selections
like these to replace the value(s).  Also, do not forget that anything can be replaced with another name (e.g., "75+").
```{r}
new_old_df[new_old_df$agegp == "75+" &
             new_old_df$alcgp == "120+" &
             new_old_df$tobgp == "10-19", ]
new_old_df[new_old_df$agegp == "75+" &
             new_old_df$alcgp == "120+" &
             new_old_df$tobgp == "10-19", "ncases"]
```

```{r, opts.label = 'hideall'}
new_old_df <- unique(new_df[, 1:3])
new_old_df$agegp <- as.character(new_old_df$agegp)
new_old_df$alcgp <- as.character(new_old_df$alcgp)
new_old_df$tobgp <- as.character(new_old_df$tobgp)
new_old_df$ncases <- 0
new_old_df$ncontrols <- 0

for (i in 1:NROW(new_df)) {
  row = new_df[i, ]
  if (row$case == "case") { 
    col <- "ncases" 
  } else {
    col <- "ncontrols"
  }
  oldval <- new_old_df[new_old_df$agegp == row$agegp &
                         new_old_df$alcgp == row$alcgp &
                         new_old_df$tobgp == row$tobgp, col]
  new_old_df[new_old_df$agegp == row$agegp &
               new_old_df$alcgp == row$alcgp &
               new_old_df$tobgp == row$tobgp, col] <- oldval + 1
}

new_old_df$agegp <- factor(new_old_df$agegp)
new_old_df$alcgp <- factor(new_old_df$alcgp)
new_old_df$tobgp <- factor(new_old_df$tobgp)
```

