---
title: "Food Prep"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```

Exercise
--------

### Slicing, Dicing, and Concatenating Vectors

This module focuses on several common data manipulation tasks: selecting, subsetting, concatinating, merging, and sorting.
Some you have already seen before, but this is a good time for review and reinforcement.  For example, when you want 
to select a part of a `vector` use square brackets, `[]`, placing between them a numeric vector that lists the elements you
want to extract.  R has a builtin `vector` that contains the capital letters: `LETTERS`.  Let's try a few of things with that 
`vector`.  Remember we are just using `LETTERS` as a stand in for a `vector` you may want to subset in some way:

```{r}
LETTERS
# what is the 10th letter?
LETTERS[10]
# what are the 15th - 22nd letters?
LETTERS[15:22]
# what are the odd indexed letters?
LETTERS[c(1,3,5,7,9,11,13,15,17,19,21,23,25)]
```

The latter is a little annoying.  That seems just like a task that a computer should be able to do for you. 
Indeed, it can, and there is a function in R called `seq` that can easily construct some complex sequences:

```{r}
# start at 1, end at 26, go by twos
seq(1, 26, by = 2)  
LETTERS[seq(1, 26, by = 2)]
# how about the even ones?
LETTERS[seq(2, 26, by = 2)]
```

Another good function: `rep`.  It is technically a mnemonic for "*rep*licate"" but I typically remember "*rep*eat":

```{r}
# repeat the number 5 10 times
rep(5, 10)
# repeat c(1, 2) 5 times
rep(c(1, 2), 5)
# repeat 1, 5 times; and 2, 5 times
rep(c(1, 2), each = 5)
# repeat 1, 3 times; 2, 2 times; 3, 1 time
rep(1:3, 3:1)
```

And who said you can't ask for something more than once when you subset a vector?  Nobody!

```{r}
LETTERS[rep(1:3, 3:1)]
```

Want to get rid of something?  Use negative numbers:

```{r}
LETTERS[-26:-21]
# another way to get the even numbers:
LETTERS[-seq(1, 26, by = 2)]
```

Logical vectors return the true ones:

```{r}
x <- 1:5
x[c(TRUE, FALSE, TRUE, TRUE, FALSE)]
# yet another way to get the even letters
# (remember R "recycles" shorter vectors)
LETTERS[c(FALSE, TRUE)]
```

Remember you can create logical vectors using the **relational** and **logical operators**.  Let's create a random 
sample of 100 letters (remember the `set.seed` is just so you can exactly replicate my example).

```{r}
set.seed(596)
letsamp <- sample(LETTERS, 100, replace = TRUE)
letsamp
letsamp == "P"
# how many Ps? add it up - remember TRUE becomes 1 when using as a numeric vector
sum(letsamp == "P")
# mind your Ps and Qs
letsamp[letsamp %in% c("P", "Q")]
```

Finally, don't forget you can use `c` to concatenate multiple vectors together, not just single numbers:
```{r}
c(LETTERS[1:3], c("H", "I"), LETTERS[24:26])
```

### Slicing, Dicing, and Concatenating Dataframes

Load in the datasets with:
```{r, eval = FALSE}
library(devtools) # install_github("advdatamgmt/adm") if you need to
```

```{r}
library(adm)
```

A lot of what you need is analogous to vectors.  Just remember that `data.frame` has rows and columns, rows come first.

```{r}
# first 6 rows of cholera_merge
cholera_merge[1:6, ]
```

Using a logical to select rows:
```{r}
cholera_merge[cholera_merge$area == 4, ]
# or the subset command
subset(cholera_merge, area == 4)
```

The second seems easier, so why use the first?  You'll need it if you want to assign back to your selection.
For example:

```{r}
cholera_merge[cholera_merge$area == 4, ]$area <- 6
# none
subset(cholera_merge, area == 4)
# because they are here
subset(cholera_merge, area == 6)
```

```{r, error = TRUE}
# doesn't work
subset(cholera_merge, area == 6)$area <- 6
```

```{r}
# put it back
cholera_merge[cholera_merge$area == 6, ]$area <- 4
```

```{r}
# women from area 3
cholera_merge[cholera_merge$sex == "F" & cholera_merge$area == 3, ]
```

Columns work the same and together with rows if you want (here I used both to keep the output short):

```{r}
cholera_merge[1:3, 3:4]
cholera_merge[1:3, c("age", "sex")]
cholera_merge[cholera_merge$area == 4, c("age", "sex")]
```

Use `cbind` to "bind" columns together and `rbind` to "bind" rows together (concatination of columns and rows).  Let's
divide up `cholera_merge` into smaller datasets.  We can use a loop and the `assign` function which has the same effect as
`<-` but works as a function (allowing us to programmatically specify the name).  The function `paste0` pastes together `character`
vectors without any space between them.

```{r}
paste0("a", "b", "c")
paste0("cholera", 1:5)
assign("x", 1:3)
x

for(i in 1:5) {
  assign(paste0("cholera", i), cholera_merge[cholera_merge$area == i, ])
}

ls() # lists your assigned variables

cholera1
cholera2
```

Now put it back together:

```{r}
c_m <- rbind(cholera1, cholera2, cholera3, cholera4, cholera5)
c_m[1:6, ]
NROW(c_m) == NROW(cholera_merge)
```

Do something strange with the columns:

```{r}
cbind(cholera1[, 1:2], cholera2[, 3:4])
```

What if you `cbind` them a few of the little datasets together?

```{r}
# extra set of parentheses prints the assignment
(threecho <- cbind(cholera1, cholera2, cholera3))
# hmmm... variables with same name
names(threecho)
# guess what you can assign directly to names
names(threecho) <- paste0(c("area", "subject", "age", "sex"), rep(1:3, each = 4))
threecho
```

### Merging dataframes

Merging is the way you combine datasets with common variable names.  Take `cholera_disease` as an example:

```{r}
cholera_disease[1:5, ]
```

It shares the `subject` with `cholera_merge` so we can merge them together:

```{r}
cmer <- merge(cholera_merge, cholera_disease)
cmer[1:10, ]
```

R automatically determines the common columns if their names are the same, but what if they are not?
You'll have to use some of the optional arguments to the function.

```{r}
names(cholera_disease) <- c("id", "disease")
cmer2 <- merge(cholera_merge, cholera_disease)
# what happened?
NROW(cmer)
NROW(cmer2)
```

What happened?  Since there were no common columns R created the merge by combining every observation in the first
`data.frame` with each observation in the second: $25 x 25 = 625$.  So, tell R what to merge on (the first dataset is 
`x` and the second is `y` in the naming convention):

```{r}
NROW(merge(cholera_merge, cholera_disease, by.x = "subject", by.y = "id"))
```

What if observations are missing in one of the `data.frames`?

```{r}
set.seed(596) # because sample is random
(cdsamp <- cholera_disease[sample(1:NROW(cholera_disease), 5), ])
merge(cholera_merge, cdsamp, by.x = "subject", by.y = "id")
```

Holy disappearing rows, Batman!  Merge only keeps the matches.  Be careful with that fact and be sure to specify another option if you 
want to keep all the rows of one:

```{r}
(merall <- merge(cholera_merge, cdsamp, by.x = "subject", by.y = "id", all.x = TRUE))
merall[20:25, ]
```

All the rows from the `x` `data.frame` are there with missing data from the `y` one represented by `NA`.  Most frequently you want
to use one or both of the all options which will save you a lot of heartache when you are merging datasets that you do not realize
are incompatible.  

### Sorting vectors and dataframes

Sorting is a common operation.

```{r}
sort(merall$age)
```

But what if you want to sort *by* something in a `data.frame`.  What you would want to know is the "subset" that was ordered in 
a way that it returned the elements in sorted order.  That's where the `order` function comes in:

```{r}
merall$age
order(merall$age)
```

See how if you take each element of `merall$age` given by `order` you will get the sorted list?  What is the 2nd element of
`merall$age`? `20` The 11th? `20`.  So:

```{r}
merall$age[order(merall$age)]
identical(merall$age[order(merall$age)], sort(merall$age))
```

`order` is more useful for `data.frames` where it can be used to sort the `data.frame`.  Here we sort `merall` by `age`:

```{r}
merall[order(merall$age), ]
```

### Extracting the first and last rows of dataframes

`head` and `tail` extract the first and last elements of a `vector` or the first and last rows of a `data.frame` (6 by default):

```{r}
head(merall)
tail(merall)
```

You can add an optional argument to tell it how many to extract:

```{r}
head(merall, 1)
tail(merall, 2)
```

You can use negative numbers to extract "all but" that many:

```{r}
head(merall, -10)
tail(merall, -15)
```


Explore and Extend
------------------

+ Write a function named `extreme` that takes a `numeric vector` as input and returns a `list` of two elements: the first named
`smallest` containing the smallest three elements in the `numeric vector` and the second `list` element named `biggest` containing
the largest three elements in the `numeric vector`.

+ Write a function `sampdf` that takes a `data.frame` as an argument and single number and returns a random sample of the rows of
the dataframe with the number of rows specified in the argument.

+ Write a loop that takes the `cholera_merge` and extracts the rows containing the oldest person in each area.  Hint: combine select/subset techniques, with `order` and `tail`.  For now, just use the `print` function to print out each row.

Evaluate
--------

Take the last "Explore and Extend" exercise one more step and use `rbind` to create a new `data.frame` from the rows.  On the
first round of the loop, you'll need to save the first row *as* the new `data.frame`.  On the subsequent **iterations** (rounds
through the loop), you'll need to `rbind` to the `data.frame` and save that to the `data.frame` variable name you choose.  Hear some
branch logic that description?

```{r, opts.label = "hideall"}
for(i in 1:5) {
  x <- cholera_merge[cholera_merge$area == i, ]
  x <- x[order(x$age), ]
  t <- tail(x, 1)
  if(i == 1) {
    df <- t
  } else {
    df <- rbind(df, t)
  }
}
```

Then, *think* about how you can use this for our `esoph` transformation from case-control to long format.  Try out your ideas as time
permits.