---
title: "Grammar of Graphics"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child = 'config/key_setup.Rmd'}
```

Exercise
--------

### Install ggplot2

Grammar of graphics is found in package `ggplot2`.  

```{r, eval = 5}
# once:
install.packages("ggplot2")

# everytime:
library(ggplot2)
``` 

### Using ggplot2

The function is called `ggplot` (note no number 2 on the end of the
function).  You want to store this output in a name because all you are doing 
below is telling ggplot how to "think"" about the data, i.e., what represents x
and y in this case.  If you run it directly, you'll get an error "No
layers in plot" because you've not added a layer to draw.  

You will need to refer frequently to the `ggplot2` documentation
found at <http://docs.ggplot2.org/current/>.  Note that the
functions often have "aesthetics" listed, that control the way the data 
is displayed in addition to the usual "arguments" that all R functions have.  
Aesthetics are placed within a function
called `aes` within your function call as an argument.  Watch
for that pattern below as we work to recreate a version of our optic nerve
head figure:

```{r, }
library(adm)
myplot <- ggplot(onhlong, aes(x = age, y = mean))
``` 

Now we can add a layer to plot the points:

```{r, tidy = FALSE}
myplot + geom_point()
``` 

#### More aesthetics

But we want the different points to have colors depending on the
groups right?  This is an aesthetic (see the documentation for
`geom_point` to see what else you can control).

```{r, tidy = FALSE}
myplot + geom_point(aes(color = case))
``` 

Yet we really want three groups right?  To do that we need a little
trick to combine the two variables into a new one.  Just paste them
together and save it as a new variable to the dataset:

```{r, tidy = FALSE}
# the trick - see what it did?  "1-Yes", "0-No", and "1-No" are three groups
(onhlong$group <- factor(with(onhlong, paste(case, clinhypo, sep = "-"))))
# need to resave myplot because it "stores" a memory of the dataset before we made
# the change
myplot <- ggplot(onhlong, aes(x = age, y = mean))
myplot + geom_point(aes(color = group))
``` 

#### Regression Lines

Good start.  Now let's add a regression line with a 95% confidence limit
based on the data itself:

```{r, tidy = FALSE}
myplot + 
    geom_point(aes(color = group)) + 
    stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"))
``` 

But you wanted the prediction interval in this case.  There is no
automatic way to create the prediction band with ggplot.  So we'll use
our `onhfit` dataset.

```{r, tidy = FALSE}
myplot +
  geom_point(aes(color = group)) +
  # se = FALSE just for the line (no standard error)
  stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"), se = FALSE) +
  # now the "ribbon"
  geom_ribbon(data = onhfit, aes(x = age, ymax = upr95, ymin = lwr95))
``` 

Ooops... order of layers and transparency matters:

```{r, tidy = FALSE}
myplot +
  geom_ribbon(data = onhfit, aes(x = age, ymax = upr95, ymin = lwr95)) +
  geom_point(aes(color = group)) +
  stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"), se = FALSE)
``` 

or
```{r, tidy = FALSE}
myplot +
  geom_ribbon(data = onhfit, aes(x = age, ymax = upr95, ymin = lwr95, alpha = 0.01)) +
  geom_point(aes(color = group)) +
  stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"), se = FALSE) 
``` 

#### Legends

Control the legends with `guides()`:

```{r, tidy = FALSE}
myplot +
  geom_ribbon(data = onhfit, aes(x = age, ymax = upr95, ymin = lwr95, alpha = 0.01)) +
  geom_point(aes(color = group)) +
  stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"), se = FALSE) +
  # just get rid of them for now; use the term that created the legend to get rid of it
  guides(color = "none", alpha = "none")
``` 


#### Axes

Control the axes with `scale_...` functions:

```{r, tidy = FALSE}
myplot + 
geom_ribbon(data = onhfit, aes(x = age, ymax = upr95, ymin = lwr95, alpha = 0.01)) +
geom_point(aes(color = group)) +
stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"), se = FALSE) +
# the specify the axis of interest and the variable type
scale_y_continuous(name = "Mean optic nerve size, mm") + 
# change the title
guides(color = "none", alpha = "none") 
``` 

Add back the color legend only and change the labels of the color "axis":

```{r, tidy = FALSE}
myplot +
geom_ribbon(data = onhfit, aes(x = age, ymax = upr95, ymin = lwr95, alpha = 0.01)) +
geom_point(aes(color = group)) +
stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"), se = FALSE) +
scale_x_continuous(name = "Age, years") +
scale_y_continuous(name = "Mean optic nerve size, mm") +     
# the labels are part of the "scale" think of the color as a third axis
scale_color_discrete(labels = c("Control", "Case, Non-ONH eye", "Case, ONH eye")) + 
# change the title
guides(color = guide_legend(aes(title = "Eye type")), alpha = "none") 
``` 

#### Themes

Make it "cleaner" for publications, etc:

```{r, tidy = FALSE}
myplot +
geom_ribbon(data = onhfit, aes(x = age, ymax = upr95, ymin = lwr95, alpha = 0.01)) +
geom_point(aes(color = group)) +
stat_smooth(method = "lm", data = subset(onhlong, group == "0-No"), se = FALSE) +
scale_x_continuous(name = "Age, years") +
scale_y_continuous(name = "Mean optic nerve size, mm") +     
scale_color_discrete(labels = c("Control", "Case, Non-ONH eye", "Case, ONH eye")) + 
guides(color = guide_legend(aes(title = "Eye type")), alpha = "none") +
# a very "clean" theme
theme_bw()
``` 


#### Pure exploration

Non-parametric smoother with 95% CI.  Big for the green band because
it is uncertain due to a lack of data:

```{r, tidy = FALSE}
myplot + 
    geom_point(aes(color = group)) + 
    geom_smooth(aes(color = group, fill = group))
``` 

Separate plots but on same axis:

```{r, tidy = FALSE}
myplot + 
    geom_point(aes(color = group)) + 
    geom_smooth(aes(color = group, fill = group)) + 
    facet_grid(~ group)
```

Explore and Extend
------------------

The `rop` dataset contains data on where three
types of physicians (variable `type`) thought the macula position
was (variables `x` and `y`) on 30
photographs (variable `picture`).  

```{r, }
# library(devtools); install_github("advdatamgmt/adm")
library(adm)
head(rop)
``` 

Can you create the following picture?
(It only uses commands and aesthetics above, you'll get some warnings
because there are some missing data points --- you can safely ignore those):

```{r, echo = FALSE, warning = FALSE}
p <- ggplot(rop, aes(x, y, color = type)) + geom_point()
p + facet_wrap(~ picture) + theme_bw()
``` 

Evaluate
--------

Modify the figure to display only the data from the retina physicians
(`type = "ret"`) and add a regression line (even though it doesn't make much sense here)
to each graph within your figure.

```{r, opts.label = "hideall", warning = FALSE}
p <- ggplot(subset(rop, type == "ret"), aes(x, y, color = type)) + 
              geom_point() + stat_smooth(method="lm")
p + facet_wrap(~ picture) + theme_bw()
``` 
