---
title: "Modeling Example"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child = 'config/key_setup.Rmd'}
```

```{r set-options, echo=FALSE, cache=FALSE}
options(width=80)
```

## Step 1. Variable Specification

I'm going to "skip" this step because you can trust all the data is
clean in this case and I want to get down to business.  I'm going to
develop a model with the outcome of interest being
whether the patient was examined by the ED physician or not and the
primary exposure of interest being the patients presenting complaint/condition.
The presenting complaint/conditions are coded as four indicator
variables because patients, as you know, can have more than one
complaint/condition: headache, neurological deficit, vision complaint , or diastolic blood pressure >120 mmHg.  The other variables of
interest (i.e., potential confounders and interacting variables)
are the examination method available (which depended on the
phase of the study), sex, race, age, body mass index, and mean arterial pressure.

## Step 2. Model Specification

The outcome is dichotomous so I'll use a logistic regression. In R, I
usually use the `rms` library.  The data is found in the `adm` package.

```{r}
library(rms)
library(adm)
head(edfun)
``` 

Is there any missing data in the variables we care about?  

```{r}
all(complete.cases(edfun))
```

No; so with that out of the way, let's now look for non-linearity:

```{r}
(m <- lrm(examined ~ cc.ha + cc.neuro + cc.vision + cc.bp + phase + sex + 
            race.black + rcs(age, 5) + rcs(bmi, 5) + rcs(map, 5), edfun))
anova(m)
``` 

Note that age is close to being non-linear and that the "total"
nonlinearity is also nearly significant.  I'm going to choose to get
rid of the non-linear term in this model to keep it "simple."  You
could choose to be extra careful about confounding control and keep
age as non-linear for now. 

Next, collinearity:

```{r}
library(perturb) # helps print with colldiag.alt - install if don't have
(m <- lrm(examined ~ cc.ha + cc.neuro + cc.vision + cc.bp + 
            phase + sex + race.black + age + bmi + map, edfun))
colldiag.alt(m)
``` 

Only mean arterial pressure and the intercept appear collinear so we
will proceed.

## Step 3. Interaction assessment

```{r}
(m <- lrm(examined ~ (cc.ha + cc.neuro + cc.vision + cc.bp) * (phase + sex + 
                                        race.black + age + bmi + map), edfun))
anova(m)
``` 

There are a couple of significant second order terms race and vision complaint
and sex and neuro complaint.  Make sure there is no evidence of
additional interaction if we reduce the model to just keep those two terms:

```{r}
(m2 <- lrm(examined ~ cc.ha + cc.neuro + cc.vision + cc.bp + phase + 
             sex + race.black + age + bmi + map + race.black * cc.vision + 
             sex * cc.neuro, edfun))
# non-significant likelihood ratio test 
# between the two models so ok to reduce
lrtest(m, m2)  
# even though the two terms are non-sig at 0.05 they are at 0.1, 
# and in addition total interaction is significant at 0.05 level now
anova(m2) 
``` 

## Step 4. Confounding assessment

Now compare the adjusted ORs for the complaints (but we have to deal
with the interactions too in this case because they effect the
exposures of interest) with the model where age, BMI, and MAP are
removed.  Kept phase because it is such an important variable.

```{r}
(m3 <- lrm(examined ~ cc.ha + cc.neuro + cc.vision + cc.bp + phase + 
             sex + race.black + race.black * cc.vision + sex * cc.neuro, edfun))
``` 

You can eyeball that the only one that is an issue is the DBP >
120 condition.  Nothing else will be confounded if we get rid age, 
BMI, and MAP.  So, now, I'm going to look just at the DBP > 120
coefficient for all subsets of age, BMI, and MAP:

```{r}
# create all subsets of age, bmi, map
# combn(list, n) creates the permutations on the rows
# when we use sapply we get to supply a different number to the function for 
# 1, 2, or 3 of the variables simultaneously
(vargroups <- sapply(1:3, function(x) combn(Cs(age, bmi, map), x)))
# next I need to run each list element through a function 
# that will create the right model based on the rows of 
# each list element; paste them together with collapse = ' + '
(formulas <- lapply(vargroups, function(m) apply(m, 2, function(x) paste(x, collapse = ' + '))))
# now make them a single vector of formulas
(formulas <- do.call(c, formulas))
# paste the remainder of the formula using paste with sep = ' + ' 
# (collapse would have made one really long formula)
# rather than 7 single formulas
baseform <- "examined ~ cc.ha + cc.neuro + cc.vision + cc.bp + cc.vision * race.black + cc.neuro * sex + phase"
(formulas <- paste(baseform, formulas, sep = ' + '))
# and don't forget the baseform is a formula too!
(formulas <- c(baseform, formulas))
# now run lrm on each of these formulas
(models <- lapply(formulas, function(form) lrm(as.formula(form), edfun)))
# no need to look at each of them; just pull out the exp of the cc.bp coef
(ors <- sapply(models, function(x) exp(coef(x)['cc.bp=Yes'])))
# match with the formulas
cbind(formulas, ors)
``` 

Note the simplest model that fully controls for confounding is the 3rd from
the bottom which requires keeping age and MAP.  So, let's plan to drop
BMI.

## Step 5. Precision

```{r}
# new function that makes it much easier to create new models:
# . means what was there before
(m4 <- update(m2, . ~ . - bmi))
``` 

Look at all the coefficients for the complaints and conditions.  They
are generally a little smaller.  No need to keep age and BMI for political
reasons so we have the final model (m4).

## Step 6. Recheck assumptions and GOF

It's all good:

```{r, error=TRUE}
# no nonlinearity at 0.05
anova(update(m4, . ~ . - age + rcs(age, 5) - map + rcs(map, 5)))
# no collinearity (only intercept and map - safe to ignore)
colldiag.alt(m4)
# model has no evidence of lack of fit
residuals(m4, "gof") # need to put x = TRUE, y = TRUE in fit for this to work (see error)
residuals(update(m4, . ~ ., x = TRUE, y = TRUE), "gof")
``` 

## Step 7. Report

Create tables...