---
title: "Rcel, part 1"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```


Exercise
--------

Once you learn R well, you will likely find that you do not tend to use spreadsheets like Microsoft Excel as much for common tasks that you 
might have previously done in such software.  We will see a spreadsheet-like structure develop as we work with the
composite data types of `vector`, `factor`, `list`, and `data.frame`.  Type the following at the R console and 
verify you get the same results, but also try to be sure you understand what is happening.  This first part will cover only
`vector` and `factor`.

### Vectors 
A `vector` is constructed using a function called simply `c`.  Note that `c` is lower case which matters because R is a *case-sensitive* language. If you type a capital, or upper case, `C` instead of lower case `c` then R will *not* know what you mean.  

```{r}
c(1,2,3)
```

A convenient way to make a vector of consecutive numbers is to use the `:` between them:

```{r}
1:5
```

Vectors can only be made of values from the same atomic type:

```{r}
c("a","b","c")
c(TRUE,FALSE,FALSE,TRUE)
```

The `is.numeric`, `is.character`, and `is.logical` functions that you learned about in [Atomic Data Types](atomic.html) work on vectors too:

```{r}
is.numeric(1:3)
is.logical(1:3)
is.character(c("a","b"))
is.logical(c(TRUE,TRUE,TRUE,FALSE,TRUE))
```

### Assignment
By now you are probably pretty tired of typing things like `c(TRUE,TRUE,TRUE,FALSE,TRUE)` and wonder if there is a way you could save things
like that for later.  There is!  It is called **assignment**.  We can **assign** a **name** to one of our vectors (or any R object).  We do this
using the `<-` operator (that is made of two characters, the less than symbol, `<`, and a hyphen, `-`).  Think of it as an arrow that directs
the value on the right hand side of `<-` into the name on the left hand side.

```{r}
x <- c(TRUE,TRUE,TRUE,FALSE,TRUE)  # nothing appears to happen, but everything is ok...
x                                  # ...see? I told you so! :)
is.logical(x)
```

The name `x` is just one of a nearly infinite number of names, but R does have some rules for valid **names**:

+ A valid **name** can only consist of:
    + letters, 
    + numbers,
    + the dot or period character (`.`), and 
    + the underscore or underline character (`_`),
+ But it can only start with:
    + a letter, or 
    + the dot *not* followed by a number. 

When you **name** something the same thing, it is replaced:
    
```{r}
x
x <- c(1,2,3)
x
```

The **name** can stand in for the object that is **assigned** to it anywhere that the object can be used, even in assignment:
```{r}
y <- x
y
```

Finally, you can not use one of the **reserved** words as **names**. You have seen a few of these already, e.g., `TRUE`, `NaN`.  It could really wreck havoc on R if you could change the value of those **names**.

### Factors

A `factor` is similar to a `vector`, but is used to represent a nominal or ordinal variable.  This allows R to automatically give you back the
right statistics for something that is not numeric.  For now, we will use the `summarize` function to show the difference.

```{r}
x <- c(1,1,2,3,1)
y <- factor(x)
summary(x)
summary(y)
```

That last output may be a little confusing at first, but notice that it is a type of table with the different values in the first row and the
number of elements in the factor that take that value.  When you give the values labels, as you often will, it can be much easier to understand
the summary of a factor.

```{r}
state <- factor(x,labels=c('GA','FL','AL'))
state
summary(state)
```

The `labels=c(...)` is a **named argument** to the `factor` function.  The **name** is `labels` and the **argument** is the `c(...)` itself.  Notice
the `=` sign that connects the **name** to the **argument**.  We will discuss this a lot more in the future.  For now, just use this exact syntax
if you need to label a `numeric vector`.

Note that the labels match the order of the values if you use numbers, not the order they appear in the `vector` you convert to a `factor`:
```{r}
x <- c(2,3,1,1,1)
state <- factor(x,labels=c('GA','FL','AL'))
state
summary(state)
```

You can also use a character `vector` to create a `factor`:
```{r}
state <- factor(c("GA","GA","FL","AL","GA"))
state
summary(state)  # notice it is the same data, but alphabetically sorted
```

### Extract elements from vectors and factors

When you want to extract a specific element from a `vector` or `factor`, you use square brackets (i.e.,`[ ]`) and the **index** (the number) of the
element within the data structure:
```{r}
x               # same x as above
x[5]
x[2]
x[2] + x[5]
x[c(2,5)]       # create a new vector with the second and fifth element of x
state[3]        # works with vectors too
```

### Operating on vectors and factors

Most operators in R operate elementwise on `vectors`, e.g.:
```{r}
x * 2
2 ^ x
x ^ 2
5 - x
x < 2
```

You can also apply operators to two `vectors` of the same length in which case the resulting `vector` will be the result of the operator
applied to the first element of each `vector`, then the second element of each `vector`, and so on:
```{r}
x - c(1,0,1,0,1)
```

You can even use a `vector` that is shorter than the other.  It will be recycled.
```{r}
c(1,0,1,0,1,0) + c(0,1)
```

However, you will get a `warning` if the shorter `vector` is not a multiple of the longer `vector` because this usually unintentional and most
often indicates a bug in your program.  Here it works the way we expect:
```{r}
c(1,0,1,0,1) + c(0,1)
```

A `warning` is issued by a program when it can continue executing, but is not sure that how it is going to continue to execute is 
what you were expecting it to do. A `warning` does not rise to the level of an `error` which is something that the program cannot recover from.

Explore and Extend
------------------

+ Make a really long vector like this: `1:100`.  Explain why R has been printing `[1]` at the beginning of the output all this time.

+ Try and assign something to the **reserved** word `TRUE`.  What is the error?  Can you make sense of it?

+ Can you guess what function tells you whether something is a `factor` or not?

+ Is a `factor` a `vector`?  Justify your answer.

+ Remember that I said that a vector can only be made from the same atomic type?  Try the following (and additional similar examples)
and try to understand what is going on.  What is R doing?  Why does this always work?
```{r}
c(TRUE,0,FALSE,3)
c("a",3,TRUE)
```

Evaluate
--------
You know the drill.  The stem of your R file should be `rcel1_`.

Before continuing add the following to your R file and execute it.  It creates random `vectors` for use in these exercises that will be the same
for everyone (more on this at a later date).

```{r}
set.seed(15437)
x <- floor(runif(100)*100)
y1 <- sample(c(TRUE,FALSE),10,replace=TRUE)
y2 <- sample(c(TRUE,FALSE),10,replace=TRUE)
z <- sample(c("a","b","c"),1000,replace=TRUE)
```

+ Is `2dogs` a valid name in R?  Why or why not?

+ What is the sum of the 11th and 82nd element of `x`?
```{r, opts.label='hideall'}
x[11] + x[82]
```

+ What is the 5th element of `y1 & y2`?
```{r, opts.label='hideall'}
y3 <- y1 & y2
y3[5]

(y1 & y2)[5]   # without creating a new variable
```

+ What is the 62nd element of `z`?
```{r, opts.label='hideall'}
z[62]
```

+ How many letter c's are in `z`?
```{r, opts.label='hideall'}
summary(factor(z))[3]   # did you know the oupput is a vector?
```