---
title: "‘Rithmatic"
author: "Beau B. Bruce"
output: 
  html_document:
      theme: readable
      highlight: tango
      css: config/styles.css
output_dir: html
params:
  key: TRUE
---

```{r key setup, child='config/key_setup.Rmd'}
```

Exercise
--------

One of the best ways to start learning to use the R console is as a calculator.  Try typing the following at the R console and verify that you get the same answers:

### Addition
```{r}
2 + 2  # This should equal 4
```

> Did you notice my use of the `#` to add some commentary to my code?  The `#` character, called a **hash** or **pound** character tells
> R to ignore the rest of the line because it was written for humans.  They are also used when displaying the output in our exercises so 
> that you could easily cut and paste a section of several continous statements without causing a problem with the output.  As you can
> see in the output we use two `#` but this is not requried as R would ignore anything after the first one.

```{r}
3 + 4 + 6
```

### Subtraction
```{r}
20 - 4.2
```

```{r}
100 - 20 - 5
```

### Multiplication
```{r}
2 * 2
```

```{r}
5 * 4 * 3 * 2 * 1
```

### Division
```{r}
25 / 5
```

```{r}
3 / 5
```


Explore and Extend
------------------

+ What do you get when you divide by zero?

+ What does the operator `%%` do?  Don't look it up - try to figure it out by experimenting with different inputs.  Here is one example of the type of things you should try:
```{r, eval=FALSE}
12 %% 5
```

+ Try to think of a situation where the operator `%%` might be useful for some procedure or process?

+ You can use parentheses to group operations when necessary, e.g. `(3 * 4) + (5 * 6)`.  In the absence of parentheses, most computer languages, including R, use a set of rules called **presidence** <a name="presidence"/> to determine which operation to do first.  Although you might think left-to-right would be the rule, this is the rule only when two operators have equal **presidence**.  Instead, certain operators take **presidence** over other operators and those with a higher **presidence** are executed first.  Try some expressions like those below and figure out the presidence of the four main mathematical operators we learned about `+`, `-`, `*`, `/`.  
    + Which have equal presidence (i.e., are evaluated left-to-right)?  
    + Which have higher presidence than the others?
```{r, eval=FALSE}
2 + 3 * 4
3 / 2 * 5
```


Evaluate
--------

Complete the following exercises in R and turn in your code (in an .R file named rithmatic_\<yourlastname\>.R) and answers (as comments within your .R file), e.g.,

```{r, eval=FALSE}
# Question 1
3 + 4 + 5
# Answer: 12

# Question 2
...
```

In a couple of weeks we will learn a better way to combine code and the answers, but for now we will rely on this method.

1. What is the result of `(2.12 + 4.2) * 1.001 / 2`?
```{r, opts.label='hideall'}
(2.12 + 4.2) * 1.001 / 2
```

2. What does the `^` operator do?  Try to figure it out without looking it up.  Give an example and describe in terms of another operator you have already learned.
```{r, opts.label='hideall'}
2 ^ 3
# ^ is the exponentiation operator; it multiplies (*) the first number by 
# itself the number oftimes given by the secord number
# so 2 ^ 3 is 2 raised to the third power or 2 * 2 * 2
```

3. What is the remainder of 6656127240 when divided by 9?
```{r, opts.label='hideall'}
6656127240 %% 9
```

4. What does the `%/%` operator do?  Again, try to figure it out without looking it up.  Only use whole numbers.  Which operator that you have learned about is this operator the complement to?
```{r, opts.label='hideall'}
# It is an operator for interger division.  It gives the number of times another number
# divides into another ignoring the remainder.  It is the complement to the modulo operator, %%.
6656127240 %/% 9
9 %/% 2
```

5. What does R say the answer to zero divided by zero is?
```{r, opts.label='hideall'}
0/0
```

